---
title: "pempek_stats_figures"
output:
  word_document: default
  pdf_document: default
date: "2026-03-10"
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This script is a general guide for making publication-quality visualizations using QIIME2-generated outputs

################################################################################ 

# Research Questions

1.  Does microbial community composition change over accumulated degree days?
2.  Does facility influence microbial diversity or community composition?
3.  Do soil and skin samples differ in their microbial diversity and community composition?

-   Your research question doesn’t need to be included in your R Markdown, but you should have a data analysis plan. It’s easy to go down rabbit holes if you don’t stick to analyses that directly address your questions. \################################################################################

################################################################################ 

# loading packages

################################################################################ 

# Load packages

```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(ggpubr)
library(vegan)
library(scales)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(emmeans)
library(rstatix)
library(forcats)
library(scales) 
```

################################################################################ 

# Metadata

################################################################################ 

# Import and clean metadata if needed

```{r}
metadata <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/03_metadata/metadata.txt")
```

################################################################################ 

# Alpha diversity

################################################################################ 

# Import alpha diversity data from QIIME2 output

```{r}
# Alpha diversity metrics
shannon <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/shannon.tsv")
evenness <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/evenness.tsv")
observed_features <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/observed_features.tsv")
faiths_pd <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/faith_pd.tsv")
```

# Tidy metrics and add to metadata

```{r}
# Shannon 
shannon <- shannon %>%
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Evenness
evenness <- evenness %>% 
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Observed Features
observed_features <- observed_features %>% 
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Faiths PD
faiths_pd <- faiths_pd %>% 
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Add to metadata
metadata <- metadata %>%
  left_join(shannon, by = "sampleid") %>% 
  left_join(evenness, by = "sampleid") %>% 
  left_join(observed_features, by = "sampleid") %>% 
  left_join(faiths_pd, by = "sampleid")
  

```

# Tidy metadata with metrics

```{r}
metadata <- metadata %>% 
  drop_na(shannon_entropy, pielou_evenness, observed_features, faith_pd)
```

# Linear Mixed Effects Models

## Time (question 1)

Shannon

```{r}
# Build model
model_age_d_shannon <- lmer(shannon_entropy ~ age_d + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_age_d_shannon)
# look at estimated variance components
summary(model_age_d_shannon)
# residual diagnostic plot, check assumptions
anova(model_age_d_shannon)
```


Evenness

```{r}
# Build model
model_day_evenness <- lmer(pielou_evenness ~ day + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_day_evenness)
# look at estimated variance components
summary(model_day_evenness)
# residual diagnostic plot, check assumptions
anova(model_day_evenness)
```


Observed Features

```{r}
# Build model
model_day_obfeatures <- lmer(observed_features ~ day + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_day_obfeatures)
# look at estimated variance components
summary(model_day_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_day_obfeatures)
```


Faiths PD

```{r}
# Build model
model_day_faithpd <- lmer(faith_pd ~ day + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_day_faithpd)
# look at estimated variance components
summary(model_day_faithpd)
# residual diagnostic plot, check assumptions
anova(model_day_faithpd)
```


## day split into 3

Shannon

```{r}
# Add metadata column 
metadata <- metadata %>%
  mutate(day_split = cut(
    day,
    breaks = 3,
    labels = c("1-5", "5-35", "35-63"),
    include.lowest = TRUE
  ))

# Stats 
# Build model
model_daysplit_shannon <- lmer(shannon_entropy ~ day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_daysplit_shannon)
# look at estimated variance components
summary(model_daysplit_shannon)
# residual diagnostic plot, check assumptions
anova(model_daysplit_shannon)
# pairwise comparisons
emmeans(model_daysplit_shannon, pairwise ~ day_split)
```



# day figure (voltility plot)

```{r}
# Shannon
day_shannon <- ggplot(metadata, aes(x = day, y = shannon_entropy, linetype = treatment)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Shannon Entropy",
    linetype = "Sample Type",
    title = "Shannon"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_shannon

# Evennes
day_evenness <- ggplot(metadata, aes(x = day, y = pielou_evenness, linetype = treatment)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Pielou's Evenness",
    linetype = "Sample Type",
    title = "Evenness"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_evenness

# Observed Features
day_obfeatures <- ggplot(metadata, aes(x = day, y = observed_features, linetype = treatment)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Observed Features",
    linetype = "Sample Type",
    title = "Observed Features"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_obfeatures

# Faith's PD
day_faithpd <- ggplot(metadata, aes(x = day, y = faith_pd, linetype = treatment)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Faith's PD",
    linetype = "Sample Type",
    title = "Faith's PD"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_faithpd

# Combine the plots
alpha_div_day <- ggarrange(day_shannon, day_evenness,
                                  day_obfeatures, day_faithpd,
                                  ncol = 2, nrow = 2,
                                  labels = c("A","B","C","D"),
                                  common.legend = TRUE,
                                  legend = "bottom")
alpha_div_day

# Save the figure
ggsave(alpha_div_day, filename = "../05_figures/alpha_div_addOc.jpeg", device = "jpeg", dpi = 300, width = 12, height = 9)

```


```{r}
model_daysplit_shannon <- lmer(shannon_entropy ~ day_split + (1|treatment), data = metadata)

pvals <- pairs(emmeans(model_daysplit_shannon, ~ day_split)) |> as.data.frame()

# make table for stat_pvalue_manual with rounded p-values
pvals_table <- data.frame(
  group1 = c("1-5","5-35","35-63"),
  group2 = c("5-35","35-63","1-5"),
  p = round(pvals$p.value, 3),           # round to 3 decimals
  y.position = max(metadata$shannon_entropy) + c(0.1,0.3,0.5)
)

# plot
day_split_shannon <- ggplot(metadata, aes(x = day_split, y = shannon_entropy)) +
  geom_boxplot() +
  stat_pvalue_manual(pvals_table, label = "p") +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"))+
  labs(x = "Age (Days)",
       y = "Shannon Entropy",
       title = "Shannon Diversity for Age (Days) Split into 3 Groups")

day_split_shannon

# Save the figure
ggsave(day_split_shannon, filename = "../05_figures/day_split_shannon.jpeg", device = "jpeg", dpi = 300, width = 8, height = 9)
```


## Day + Treatment (question 2)

Shannon

```{r}
# Build model
model_age_treatment_shannon <- lmer(shannon_entropy ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_age_treatment_shannon)
# look at estimated variance components
summary(model_age_treatment_shannon)
# residual diagnostic plot, check assumptions
anova(model_age_treatment_shannon)
# pairwise comparisons
emmeans(model_age_treatment_shannon, pairwise ~ treatment|day_split)
```


Really quick lets just look when we don't split day

```{r}
# Build model
model_day_treatment_shannon2 <- lmer(shannon_entropy ~ treatment*day + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_shannon2)
# look at estimated variance components
summary(model_day_treatment_shannon2)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_shannon2)
# pairwise comparisons
emmeans(model_day_treatment_shannon2, pairwise ~ treatment)
```


Evenness

```{r}
# Build model
model_day_treatment_evenness <- lmer(pielou_evenness ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_evenness)
# look at estimated variance components
summary(model_day_treatment_evenness)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_evenness)
# pairwise comparisons
emmeans(model_day_treatment_evenness, pairwise ~ treatment|day_split)
```


Observed Features

```{r}
# Build model
model_day_treatment_obfeatures <- lmer(observed_features ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_obfeatures)
# look at estimated variance components
summary(model_day_treatment_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_obfeatures)
# pairwise comparisons
emmeans(model_day_treatment_obfeatures, pairwise ~ treatment|day_split)
```


Faith's PD

```{r}
# Build model
model_day_treatment_faithpd <- lmer(faith_pd ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_faithpd)
# look at estimated variance components
summary(model_day_treatment_faithpd)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_faithpd)
# pairwise comparisons
emmeans(model_day_treatment_faithpd, pairwise ~ treatment|day_split)
```


# treatment figure ajusted for time (box plot) (3 timepoint box plot)

```{r}
model_day_treatment_faithpd <- lmer(faith_pd ~ treatment*day_split + (1|calf_id), data = metadata)  
pvals <- pairs(
  emmeans(model_day_treatment_faithpd, ~ treatment | day_split)
) |> as.data.frame()
pvals_table <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  dplyr::mutate(
    p = round(p.value, 4),
    y.position = max(metadata$faith_pd) + 0.1
  )
pvals_table

```


## day + facility + sex (question 3)

Shannon

```{r}
# Build model
model_treatment_day_shannon <- lmer(shannon_entropy ~ treatment*day + sex + (1|calf_id), data = metadata)

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_shannon)
# look at estimated variance components
summary(model_treatment_day_shannon)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_shannon)
# pairwise comparisons
emmeans(model_treatment_day_shannon, pairwise ~ treatment|day)
```


Evenness

```{r}
# Build model
model_treatment_day_evenness <- lmer(pielou_evenness ~ treatment*day + sex + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_evenness)
# look at estimated variance components
summary(model_treatment_day_evenness)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_evenness)
# pairwise comparisons
emmeans(model_treatment_day_evenness, pairwise ~ treatment|day)
```


Observed features

```{r}
# Build model
model_treatment_day_obfeatures <- lmer(observed_features ~ treatment*day + sex + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_obfeatures)
# look at estimated variance components
summary(model_treatment_day_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_obfeatures)
# pairwise comparisons
emmeans(model_treatment_day_obfeatures, pairwise ~ treatment|day)
```


Faith's PD

```{r}
# Build model
model_treatment_day_faithpd <- lmer(faith_pd ~ treatment*day + sex + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_faithpd)
# look at estimated variance components
summary(model_treatment_day_faithpd)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_faithpd)
# pairwise comparisons
emmeans(model_treatment_day_faithpd, pairwise ~ treatment|day)
```

## Sample type figure

```{r}
model_treatment_day_obfeatures <- lmer(observed_features ~ treatment*day + sex + (1|calf_id), data = metadata)  

pvals <- pairs(
  emmeans(model_treatment_day_obfeatures, ~ treatment | day)
) |> as.data.frame()


pvals_table <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  dplyr::mutate(
    p = round(p.value, 4),
    y.position = max(metadata$observed_features) + 0.1
  )
pvals_table
```