---
title: "Untitled"
author: "Dominikus Atmaka"
date: "2026-05-04"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(ggpubr)
library(vegan)
library(scales)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(emmeans)
library(rstatix)
library(forcats)
library(scales) 
```

```{r}
metadata <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/03_metadata/metadata.txt")
as.character(metadata$sample_name)
metadata$sample_name <- as.character(metadata$sample_name)
```

################################################################################ 

# Beta diversity

################################################################################ 

# Load in beta diversity metrics and tidy to add to metadata

(remember this is already our PCOA results so this is already our ordinations).
I make separate metadata's for each beta diversity metric, another option would be to rename each PC ex.
PC1J for jaccard and add PC1-PC3 for each metric to the metadata.
I find it easier to keep everything separate when the metrics have different metadata.
The variances extracted from the unzipped PCoA files will not match what you see in the Emperor plot, because Emperor adjusts the variance for plotting.
To obtain variances that match the Emperor plot, you need to use qiime2R.

```{r}
# Read in beta div metrics 
bray_curtis <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/bray_curtis.txt")
jaccard <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/jaccard.txt")
uw_unifrac <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/unweighted_unifrac.txt")
w_unifrac <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/weighted_unifrac.txt")

# Clean metrics
# bray curtis
# get variance
bray_curtis_var <- bray_curtis %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# clean ordination
bray_curtis <- bray_curtis %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_bc <- bray_curtis %>%
  select(sample_name,PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Jaccard
# get variance
jaccard_var <- jaccard %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Clean ordinace
jaccard <- jaccard %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_jaccard <- jaccard %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Unweighted Unifrac
# Get variance
uw_var <- uw_unifrac %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Clean ordinance
uw_unifrac <- uw_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_uwunifrac <- uw_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Weighted Unifrac
# get variance
wunifrac_var <- w_unifrac %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Tidy ordinace
w_unifrac <- w_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_wunifrac <- w_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

```

# Bray Curtis

#PC1

```{r}
# Build model
model_day_treatment_bc1 <- lmer(PC1 ~ treatment*day + treatment + (1|calf_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_bc1)
# look at estimated variance components
summary(model_day_treatment_bc1)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_bc1)
# pairwise comparisons
emmeans(model_day_treatment_bc1, pairwise ~ treatment|day)
```

#PC2

```{r}
# Build model
model_day_treatment_bc2 <- lmer(PC2 ~ treatment*day + treatment + (1|calf_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_bc2)
# look at estimated variance components
summary(model_day_treatment_bc2)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_bc2)
# pairwise comparisons
emmeans(model_day_treatment_bc2, pairwise ~ treatment|day)
```

#PC3

```{r}
# Build model
model_day_treatment_bc3 <- lmer(PC3 ~ treatment*day + treatment + (1|calf_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_bc3)
# look at estimated variance components
summary(model_day_treatment_bc3)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_bc3)
# pairwise comparisons
emmeans(model_day_treatment_bc3, pairwise ~ treatment|day)
```

# Jaccard

PC1

```{r}
# Build model
model_day_treatment_j1 <- lmer(PC1 ~ treatment*day + treatment + (1|calf_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_j1)
# look at estimated variance components
summary(model_day_treatment_j1)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_j1)
# pairwise comparisons
emmeans(model_day_treatment_j1, pairwise ~ treatment|day)
```

PC2

```{r}
# Build model
model_day_treatment_j2 <- lmer(PC2 ~ treatment*day + treatment + (1|calf_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_j2)
# look at estimated variance components
summary(model_day_treatment_j2)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_j2)
# pairwise comparisons
emmeans(model_day_treatment_j2, pairwise ~ treatment|day)
```

PC3

```{r}
# Build model
model_day_treatment_j3 <- lmer(PC3 ~ treatment*day + treatment + (1|calf_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_j3)
# look at estimated variance components
summary(model_day_treatment_j3)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_j3)
# pairwise comparisons
emmeans(model_day_treatment_j3, pairwise ~ treatment|day)
```

# Unweighted Unifrac

PC1

```{r}
# Build model
model_day_treatment_uwunifrac1 <- lmer(PC1 ~ treatment*day + treatment + (1|calf_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_uwunifrac1)
# look at estimated variance components
summary(model_day_treatment_uwunifrac1)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_uwunifrac1)
# pairwise comparisons
emmeans(model_day_treatment_uwunifrac1, pairwise ~ treatment|day)
```

PC2

```{r}
# Build model
model_day_treatment_uwunifrac2 <- lmer(PC2 ~ treatment*day + treatment + (1|calf_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_uwunifrac2)
# look at estimated variance components
summary(model_day_treatment_uwunifrac2)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_uwunifrac2)
# pairwise comparisons
emmeans(model_day_treatment_uwunifrac2, pairwise ~ treatment|day)
```

PC3

```{r}
# Build model
model_day_treatment_uwunifrac3 <- lmer(PC3 ~ treatment*day + treatment + (1|calf_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_uwunifrac3)
# look at estimated variance components
summary(model_day_treatment_uwunifrac3)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_uwunifrac3)
# pairwise comparisons
emmeans(model_day_treatment_uwunifrac3, pairwise ~ treatment|day)
```

# Weighted Unifrac

PC1

```{r}
# Build model
model_day_treatment_wunifrac1 <- lmer(PC1 ~ treatment*day + treatment + (1|calf_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_wunifrac1)
# look at estimated variance components
summary(model_day_treatment_wunifrac1)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_wunifrac1)
# pairwise comparisons
emmeans(model_day_treatment_wunifrac1, pairwise ~ treatment|day)
```

PC2

```{r}
# Build model
model_day_treatment_wunifrac2 <- lmer(PC2 ~ treatment*day + treatment + (1|calf_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_wunifrac2)
# look at estimated variance components
summary(model_day_treatment_wunifrac2)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_wunifrac2)
# pairwise comparisons
emmeans(model_day_treatment_wunifrac2, pairwise ~ treatment|day)
```

PC3

```{r}
# Build model
model_day_treatment_wunifrac3 <- lmer(PC3 ~ treatment*day + treatment + (1|calf_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_wunifrac3)
# look at estimated variance components
summary(model_day_treatment_wunifrac3)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_wunifrac3)
# pairwise comparisons
emmeans(model_day_treatment_wunifrac3, pairwise ~ treatment|day)
```