---
title: "pempek_stats_figures"
output: pdf_document
date: "2026-03-10"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This script is a general guide for making publication-quality visualizations using QIIME2-generated outputs

################################################################################ 

# Research Questions

1.  Does microbial community composition change over accumulated degree days?
2.  Does facility influence microbial diversity or community composition?
3.  Do soil and skin samples differ in their microbial diversity and community composition?

-   Your research question doesn’t need to be included in your R Markdown, but you should have a data analysis plan. It’s easy to go down rabbit holes if you don’t stick to analyses that directly address your questions. \################################################################################

################################################################################ 

# Installing and loading packages

################################################################################ 

# Install packages

-   If you have already installed the packages then you don't need to rerun this.

```{r}
# Each of the packages we will use are CRAN packages
packages_to_install <- c("tidyverse", "ggplot2","readr","ggpubr","vegan","scales","lme4", "lmerTest","pbkrtest","emmeans", "rstatix", "forcats", "scales")

install.packages(packages_to_install)
```

# Load packages

```{r}
library(tidyverse)
library(ggplot2)
library(readr)
library(ggpubr)
library(vegan)
library(scales)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(emmeans)
library(rstatix)
library(forcats)
library(scales) 
```

################################################################################ 

# Metadata

################################################################################ 

# Import and clean metadata if needed

```{r}
metadata <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/03_metadata/metadata.txt")
```

################################################################################ 

# Alpha diversity

################################################################################ 

# Import alpha diversity data from QIIME2 output

```{r}
# Alpha diversity metrics
shannon <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/shannon.tsv")
evenness <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/evenness.tsv")
observed_features <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/observed_features.tsv")
faiths_pd <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/alpha_div/faith_pd.tsv")
```

# Tidy metrics and add to metadata

```{r}
# Shannon 
shannon <- shannon %>%
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Evenness
evenness <- evenness %>% 
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Observed Features
observed_features <- observed_features %>% 
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Faiths PD
faiths_pd <- faiths_pd %>% 
  rename(sampleid = 1)  #rename 1st column to to match metadata

# Add to metadata
metadata <- metadata %>%
  left_join(shannon, by = "sampleid") %>% 
  left_join(evenness, by = "sampleid") %>% 
  left_join(observed_features, by = "sampleid") %>% 
  left_join(faiths_pd, by = "sampleid")
  

```

# Tidy metadata with metrics

```{r}
metadata <- metadata %>% 
  drop_na(shannon_entropy, pielou_evenness, observed_features, faith_pd)
```

# Linear Mixed Effects Models

## Time (question 1)

Shannon

```{r}
# Build model
model_age_d_shannon <- lmer(shannon_entropy ~ age_d + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_age_d_shannon)
# look at estimated variance components
summary(model_age_d_shannon)
# residual diagnostic plot, check assumptions
anova(model_age_d_shannon)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```

Evenness

```{r}
# Build model
model_day_evenness <- lmer(pielou_evenness ~ day + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_day_evenness)
# look at estimated variance components
summary(model_day_evenness)
# residual diagnostic plot, check assumptions
anova(model_day_evenness)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```

Observed Features

```{r}
# Build model
model_day_obfeatures <- lmer(observed_features ~ day + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_day_obfeatures)
# look at estimated variance components
summary(model_day_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_day_obfeatures)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```

Faiths PD

```{r}
# Build model
model_day_faithpd <- lmer(faith_pd ~ day + (1|calf_id), data = metadata) # I name all my models with the variable I am testing and the metric I am testing. 

# Residual diagnostic plot, check assumptions
plot(model_day_faithpd)
# look at estimated variance components
summary(model_day_faithpd)
# residual diagnostic plot, check assumptions
anova(model_day_faithpd)
# pairwise comparisons, there are a lot of time points for this data set so this line doesn't work yet, I'll show you how to use it a little later
#emmeans(model_add0_shannon, pairwise ~ add_0c)
```

## day split into 3

Shannon

```{r}
# Add metadata column 
metadata <- metadata %>%
  mutate(day_split = cut(
    day,
    breaks = 3,
    labels = c("1-5", "5-35", "35-63"),
    include.lowest = TRUE
  ))

# Stats 
# Build model
model_daysplit_shannon <- lmer(shannon_entropy ~ day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_daysplit_shannon)
# look at estimated variance components
summary(model_daysplit_shannon)
# residual diagnostic plot, check assumptions
anova(model_daysplit_shannon)
# pairwise comparisons
emmeans(model_daysplit_shannon, pairwise ~ day_split)
```


# day figure (voltility plot)

```{r}
# Shannon
day_shannon <- ggplot(metadata, aes(x = day, y = shannon_entropy, linetype = treatment)) +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Shannon Entropy",
    linetype = "Sample Type",
    title = "Shannon"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_shannon

# Evennes
day_evenness <- ggplot(metadata, aes(x = day, y = pielou_evenness, linetype = treatment)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Pielou's Evenness",
    linetype = "Sample Type",
    title = "Evenness"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_evenness

# Observed Features
day_obfeatures <- ggplot(metadata, aes(x = day, y = observed_features, linetype = treatment)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Observed Features",
    linetype = "Sample Type",
    title = "Observed Features"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_obfeatures

# Faith's PD
day_faithpd <- ggplot(metadata, aes(x = day, y = faith_pd, linetype = treatment)) +
  geom_line(size = 0.7, alpha = 0.8) +
  theme_bw(base_size = 14) +
  labs(
    x = "Age (Days)",
    y = "Faith's PD",
    linetype = "Sample Type",
    title = "Faith's PD"
  ) +
  theme(
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.title = element_text(face = "bold"),
    legend.position = "bottom",
    legend.background = element_rect(fill = "transparent"),
    plot.title = element_text(face = "bold")
  )

day_faithpd

# Combine the plots
alpha_div_day <- ggarrange(day_shannon, day_evenness,
                                  day_obfeatures, day_faithpd,
                                  ncol = 2, nrow = 2,
                                  labels = c("A","B","C","D"),
                                  common.legend = TRUE,
                                  legend = "bottom")
alpha_div_day

# Save the figure
ggsave(alpha_div_day, filename = "../05_figures/alpha_div_addOc.jpeg", device = "jpeg", dpi = 300, width = 12, height = 9)

```

```{r}
model_daysplit_shannon <- lmer(shannon_entropy ~ day_split + (1|treatment), data = metadata)

pvals <- pairs(emmeans(model_daysplit_shannon, ~ day_split)) |> as.data.frame()

# make table for stat_pvalue_manual with rounded p-values
pvals_table <- data.frame(
  group1 = c("1-5","5-35","35-63"),
  group2 = c("5-35","35-63","1-5"),
  p = round(pvals$p.value, 3),           # round to 3 decimals
  y.position = max(metadata$shannon_entropy) + c(0.1,0.3,0.5)
)

# plot
day_split_shannon <- ggplot(metadata, aes(x = day_split, y = shannon_entropy)) +
  geom_boxplot() +
  stat_pvalue_manual(pvals_table, label = "p") +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"))+
  labs(x = "Age (Days)",
       y = "Shannon Entropy",
       title = "Shannon Diversity for Age (Days) Split into 3 Groups")

day_split_shannon

# Save the figure
ggsave(day_split_shannon, filename = "../05_figures/day_split_shannon.jpeg", device = "jpeg", dpi = 300, width = 8, height = 9)
```

## Day + Treatment (question 2)

Shannon

```{r}
# Build model
model_age_treatment_shannon <- lmer(shannon_entropy ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_age_treatment_shannon)
# look at estimated variance components
summary(model_age_treatment_shannon)
# residual diagnostic plot, check assumptions
anova(model_age_treatment_shannon)
# pairwise comparisons
emmeans(model_age_treatment_shannon, pairwise ~ treatment|day_split)
```

Really quick lets just look when we don't split day

```{r}
# Build model
model_day_treatment_shannon2 <- lmer(shannon_entropy ~ treatment*day + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_shannon2)
# look at estimated variance components
summary(model_day_treatment_shannon2)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_shannon2)
# pairwise comparisons
emmeans(model_day_treatment_shannon2, pairwise ~ treatment)
```

Evenness

```{r}
# Build model
model_day_treatment_evenness <- lmer(pielou_evenness ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_evenness)
# look at estimated variance components
summary(model_day_treatment_evenness)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_evenness)
# pairwise comparisons
emmeans(model_day_treatment_evenness, pairwise ~ treatment|day_split)
```

Observed Features

```{r}
# Build model
model_day_treatment_obfeatures <- lmer(observed_features ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_obfeatures)
# look at estimated variance components
summary(model_day_treatment_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_obfeatures)
# pairwise comparisons
emmeans(model_day_treatment_obfeatures, pairwise ~ treatment|day_split)
```

Faith's PD

```{r}
# Build model
model_day_treatment_faithpd <- lmer(faith_pd ~ treatment*day_split + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_day_treatment_faithpd)
# look at estimated variance components
summary(model_day_treatment_faithpd)
# residual diagnostic plot, check assumptions
anova(model_day_treatment_faithpd)
# pairwise comparisons
emmeans(model_day_treatment_faithpd, pairwise ~ treatment|day_split)
```

# treatment figure ajusted for time (box plot) (3 timepoint box plot)

```{r}
model_day_treatment_faithpd <- lmer(faith_pd ~ treatment*day_split + (1|calf_id), data = metadata)  
pvals <- pairs(
  emmeans(model_day_treatment_faithpd, ~ treatment | day_split)
) |> as.data.frame()
pvals_table <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  dplyr::mutate(
    p = round(p.value, 4),
    y.position = max(metadata$faith_pd) + 0.1
  )
pvals_table

#only plot the significant associations
pvals_table_significant <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  filter(p.value < 0.05) %>%                     
  mutate(p = round(p.value, 4))
pvals_table_significant

# fix overlapping significance bars
pvals_table_significant <- pvals_table_significant %>%
  add_y_position(
    data = metadata,
    formula = faith_pd ~ treatment,
    fun = "max",
    step.increase = 0.12
  )
pvals_table_significant

pvals_table_significant <- pvals %>%
  tidyr::separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  dplyr::filter(p.value < 0.05) %>%
  dplyr::mutate(p = round(p.value, 4))

max_vals <- metadata %>%
  dplyr::group_by(day_split) %>%
  dplyr::summarise(y = max(faith_pd, na.rm = TRUE))

pvals_table_significant <- pvals_table_significant %>%
  dplyr::left_join(max_vals, by = "day_split") %>%
  dplyr::mutate(y.position = y + 0.1)

colnames(pvals_table_significant)



# Plot Faiths PD per facility by the grouped ADDs, add only significant comparisons
day_split_faithpd_treatment <- ggplot(metadata,
       aes(x = treatment, y = faith_pd, fill = treatment)) +
  geom_boxplot() +
  facet_wrap(~ day_split) +
  stat_pvalue_manual(
    pvals_table_significant,
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    label = "p",
    tip.length = 0.01,
    size = 2,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = c(
    "pair" = "#ff9896",
    "individual" = "#8c6d31"
  )) +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"), 
    legend.position = "none"
  ) +
  labs(
    x = "Treatment",
    y = "Faiths Phylogenetic Diversity",
    title = "Faith PD Across Facilities Within Time Groups"
  )

day_split_faithpd_treatment
ggsave(day_split_faithpd_treatment, filename = "../05_figures/add_split_faiths_facility.jpeg", device = "jpeg", dpi = 300, width = 8, height = 9)

```

## day + facility + sex (question 3)

Shannon

```{r}
# Build model
model_treatment_day_shannon <- lmer(shannon_entropy ~ treatment*day + sex + (1|calf_id), data = metadata)

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_shannon)
# look at estimated variance components
summary(model_treatment_day_shannon)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_shannon)
# pairwise comparisons
emmeans(model_treatment_day_shannon, pairwise ~ treatment|day)
```

Evenness

```{r}
# Build model
model_treatment_day_evenness <- lmer(pielou_evenness ~ treatment*day + sex + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_evenness)
# look at estimated variance components
summary(model_treatment_day_evenness)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_evenness)
# pairwise comparisons
emmeans(model_treatment_day_evenness, pairwise ~ treatment|day)
```

Observed features

```{r}
# Build model
model_treatment_day_obfeatures <- lmer(observed_features ~ treatment*day + sex + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_obfeatures)
# look at estimated variance components
summary(model_treatment_day_obfeatures)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_obfeatures)
# pairwise comparisons
emmeans(model_treatment_day_obfeatures, pairwise ~ treatment|day)
```

Faith's PD

```{r}
# Build model
model_treatment_day_faithpd <- lmer(faith_pd ~ treatment*day + sex + (1|calf_id), data = metadata)  

# Residual diagnostic plot, check assumptions
plot(model_treatment_day_faithpd)
# look at estimated variance components
summary(model_treatment_day_faithpd)
# residual diagnostic plot, check assumptions
anova(model_treatment_day_faithpd)
# pairwise comparisons
emmeans(model_treatment_day_faithpd, pairwise ~ treatment|day)
```

## Sample type figure

```{r}
model_treatment_day_obfeatures <- lmer(observed_features ~ treatment*day + sex + (1|calf_id), data = metadata)  

pvals <- pairs(
  emmeans(model_treatment_day_obfeatures, ~ treatment | day)
) |> as.data.frame()


pvals_table <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  dplyr::mutate(
    p = round(p.value, 4),
    y.position = max(metadata$observed_features) + 0.1
  )
pvals_table

#only plot the significant associations
pvals_table_significant <- pvals %>%
  separate(contrast, into = c("group1","group2"), sep = " - ") %>%
  filter(p.value < 0.05) %>%                     
  mutate(p = round(p.value, 10))
pvals_table_significant

# fix overlapping significance bars
pvals_table_significant <- pvals_table_significant %>%
  add_y_position(
    data = metadata,
    formula = observed_features ~ treatment,
    fun = "max",
    step.increase = 0.12
  )
pvals_table_significant


# Plot Observed features from skin versus soils add only significant comparisons
day_obsFeats_treatment <- ggplot(metadata,
       aes(x = sample_type, y = observed_features, fill = treatment)) +
  geom_boxplot() +
  facet_wrap(~ day) +
  stat_pvalue_manual(
    pvals_table_significant,
    xmin = "group1",
    xmax = "group2",
    label = "p",
    tip.length = 0.01,
    size = 3
  ) +
  scale_fill_manual(values = c(
    "individual" = "#9e003f",
    "pair" = "#b15928")) +
  theme_bw() +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"), 
    legend.position = "none"
  ) +
  labs(
    x = "Treatment",
    y = "Observed Features",
    title = "Observed Features Within Day Groups",
    fill = "Treatment"
  )


day_obsFeats_treatment
ggsave(add_split_obsFeats_sampleType, filename = "../05_figures/add_split_obsFeats_sampleType.jpeg", device = "jpeg", dpi = 300, width = 8, height = 9)
```

################################################################################ 

# Beta diversity

################################################################################ 

# Load in beta diversity metrics and tidy to add to metadata

(remember this is already our PCOA results so this is already our ordinations).
I make separate metadata's for each beta diversity metric, another option would be to rename each PC ex.
PC1J for jaccard and add PC1-PC3 for each metric to the metadata.
I find it easier to keep everything separate when the metrics have different metadata.
The variances extracted from the unzipped PCoA files will not match what you see in the Emperor plot, because Emperor adjusts the variance for plotting.
To obtain variances that match the Emperor plot, you need to use qiime2R.

```{r}
# Read in beta div metrics 
bray_curtis <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/bray_curtis.txt")
jaccard <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/jaccard.txt")
uw_unifrac <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/unweighted_unifrac.txt")
w_unifrac <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/beta_div/weighted_unifrac.txt")

# Clean metrics
# bray curtis
# get variance
bray_curtis_var <- bray_curtis %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# clean ordination
bray_curtis <- bray_curtis %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_bc <- bray_curtis %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Jaccard
# get variance
jaccard_var <- jaccard %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Clean ordinace
jaccard <- jaccard %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_jaccard <- jaccard %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Unweighted Unifrac
# Get variance
uw_var <- uw_unifrac %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Clean ordinance
uw_unifrac <- uw_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_uwunifrac <- uw_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

# Weighted Unifrac
# get variance
wunifrac_var <- w_unifrac %>%
  slice(3) %>%   # row 3 contains the numeric proportion explained values
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%  # split tab-separated values
  mutate(across(starts_with("PC"), ~ as.numeric(.x) * 100)) %>%  # convert to percent
  select(PC1, PC2, PC3) 

# Tidy ordinace
w_unifrac <- w_unifrac %>%
  slice(6:n()) %>%
  filter(!Eigvals %in% c("Biplot", "Site constraints")) %>%
  separate(2, into = paste0("PC", 1:10), sep = "\t") %>%
  rename(sample_name = Eigvals) #change eigenvals to match the sampleid, sample_name etc in your metadata

metadata_wunifrac <- w_unifrac %>%
  select(sample_name, PC1, PC2, PC3) %>%
  left_join(metadata) %>% 
  mutate(across(c(PC1, PC2, PC3), as.numeric))

```

# Bray Curtis

#PC1

```{r}
# Build model
model_add0_facility_sampletype_bc1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc1)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc1, pairwise ~ sample_type|add0c_split)
```

#PC2

```{r}
# Build model
model_add0_facility_sampletype_bc2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc2)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc2, pairwise ~ sample_type|add0c_split)
```

#PC3

```{r}
# Build model
model_add0_facility_sampletype_bc3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_bc)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_bc3)
# look at estimated variance components
summary(model_add0_facility_sampletype_bc3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_bc3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_bc3, pairwise ~ sample_type|add0c_split)
```

# Jaccard

PC1

```{r}
# Build model
model_add0_facility_sampletype_j1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j1)
# look at estimated variance components
summary(model_add0_facility_sampletype_j1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j1, pairwise ~ sample_type|add0c_split)
```

PC2

```{r}
# Build model
model_add0_facility_sampletype_j2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j2)
# look at estimated variance components
summary(model_add0_facility_sampletype_j2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j2, pairwise ~ sample_type|add0c_split)
```

PC3

```{r}
# Build model
model_add0_facility_sampletype_j3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_jaccard)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_j3)
# look at estimated variance components
summary(model_add0_facility_sampletype_j3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_j3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_j3, pairwise ~ sample_type|add0c_split)
```

# Unweighted Unifrac

PC1

```{r}
# Build model
model_add0_facility_sampletype_uwunifrac1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac1)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac1, pairwise ~ sample_type|add0c_split)
```

PC2

```{r}
# Build model
model_add0_facility_sampletype_uwunifrac2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac2)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac2, pairwise ~ sample_type|add0c_split)
```

PC3

```{r}
# Build model
model_add0_facility_sampletype_uwunifrac3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_uwunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_uwunifrac3)
# look at estimated variance components
summary(model_add0_facility_sampletype_uwunifrac3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_uwunifrac3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_uwunifrac3, pairwise ~ sample_type|add0c_split)
```

# Weighted Unifrac

PC1

```{r}
# Build model
model_add0_facility_sampletype_wunifrac1 <- lmer(PC1 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac1)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac1)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac1)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac1, pairwise ~ sample_type|add0c_split)
```

PC2

```{r}
# Build model
model_add0_facility_sampletype_wunifrac2 <- lmer(PC2 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac2)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac2)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac2)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac2, pairwise ~ sample_type|add0c_split)
```

PC3

```{r}
# Build model
model_add0_facility_sampletype_wunifrac3 <- lmer(PC3 ~ sample_type*add0c_split + facility + (1|host_subject_id), data = metadata_wunifrac)  
# I always add which PC I am testing into the model name

# Residual diagnostic plot, check assumptions
plot(model_add0_facility_sampletype_wunifrac3)
# look at estimated variance components
summary(model_add0_facility_sampletype_wunifrac3)
# residual diagnostic plot, check assumptions
anova(model_add0_facility_sampletype_wunifrac3)
# pairwise comparisons
emmeans(model_add0_facility_sampletype_wunifrac3, pairwise ~ sample_type|add0c_split)
```

# Figure

```{r}
# Bray Curtis

# Get min values for placing the p-value label in bottom-left
x_min <- min(metadata_bc$PC1, na.rm = TRUE)
y_min <- min(metadata_bc$PC2, na.rm = TRUE)

# PCoA plot with sample_type coloring, ellipses, and p-value in bottom-left
sampletype_bray_curtis <- ggplot(metadata_bc, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(bray_curtis_var[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(bray_curtis_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Bray-Curtis PCoA") +
  annotate(
    "text",
    x = x_min * 1.15,             
    y = y_min * 1.15,             
    label = "LME; p = 6.618e-12",      # your model p-value, take this from the anova(model_add0_facility_sampletype_bc1) table
    hjust = 0,                     
    vjust = 0,                     
    size = 4,
    fontface = "italic"
  ) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold")
  )

sampletype_bray_curtis

# Jaccard
sampletype_jaccard <- ggplot(metadata_jaccard, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  theme_bw() + 
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(jaccard_var[1], 1), "%)")) + 
  ylab(paste0("PC2 (", round(jaccard_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Jaccard PCoA") 
# uncomment this section do add a p-value you calculated from your lme
  # annotate("text",
#            x = x_max * 1.2,
#            y = y_max * 1.2,
#            label = "p = 0.0078",
#            hjust = 1, vjust = 1,
#            size = 5, fontface = "italic") +
#   coord_cartesian(clip = "off")

sampletype_jaccard

# Unweighted Unifrac
sampletype_uwunifrac <- ggplot(metadata_uwunifrac, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  theme_bw() + 
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(uw_var[1], 1), "%)")) + 
  ylab(paste0("PC2 (", round(uw_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Unweighted Unifrac PCoA") 
# uncomment this section do add a p-value you calculated from your lme
  # annotate("text",
#            x = x_max * 1.2,
#            y = y_max * 1.2,
#            label = "p = 0.0078",
#            hjust = 1, vjust = 1,
#            size = 5, fontface = "italic") +
#   coord_cartesian(clip = "off")

sampletype_uwunifrac

# Weighted Unifrac 
sampletype_wunifrac <- ggplot(metadata_wunifrac, aes(x = PC1, y = PC2, color = sample_type)) + 
  geom_point(size = 3) +
  theme_bw() + 
  stat_ellipse(type = "t", level = 0.95) +
  xlab(paste0("PC1 (", round(wunifrac_var[1], 1), "%)")) + 
  ylab(paste0("PC2 (", round(wunifrac_var[2], 1), "%)")) +
  scale_color_manual(
    values = c("skin" = "#9e003f", "soil" = "#b15928"),
    name = "Sample Type"
  ) +
  labs(title = "Weighted Unifrac PCoA") 
# uncomment this section do add a p-value you calculated from your lme
  # annotate("text",
#            x = x_max * 1.2,
#            y = y_max * 1.2,
#            label = "p = 0.0078",
#            hjust = 1, vjust = 1,
#            size = 5, fontface = "italic") +
#   coord_cartesian(clip = "off")

sampletype_wunifrac


# Combine metrics plots
beta_div_sample_type <- ggarrange(sampletype_bray_curtis, sampletype_jaccard,
                                  sampletype_uwunifrac, sampletype_wunifrac,
                                  ncol = 2, nrow = 2,
                                  labels = c("A","B","C","D"),
                                  common.legend = TRUE,
                                  legend = "bottom")
beta_div_sample_type

# Save the figure
ggsave(beta_div_sample_type, filename = "../05_figures/beta_div_sample_type.jpeg", device = "jpeg", dpi = 300, width = 10, height = 9)
```

################################################################################ 

# Taxa bar plots

################################################################################ 

# Import feature table and taxonomy data from QIIME2 output

```{r}
tabulated_results <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/04_code/taxonomy/tabulated_results.tsv")
```

# Format, convert taxonomy to relative abundance, and group at taxonomic levels

```{r}
# format data for making a taxa plot
# get rid of #q2:types row
tabulated_results <- tabulated_results[2:nrow(tabulated_results),]

# remove non-numeric columns
# we will filter out Taxon for now so we can calculate relative abundances, but will add it back later
tabulated_results_filt <- tabulated_results %>% 
  select(-all_of(c("Sequence","Confidence","Taxon")))
# we keep id and make it as the row name
tabulated_results_filt <- column_to_rownames(tabulated_results_filt, var = "id")

# calculate relative abundances
tabulated_results_filt[] <- lapply(tabulated_results_filt, function(x) as.numeric(as.character(x)))
rel_abundance <- apply(tabulated_results_filt, 2, function(x) x / sum(x))

# check that each column adds up to 1
colSums(rel_abundance)

# re-add taxonomy to relative abundance table
rel_abundance <- data.frame(rel_abundance, check.names = FALSE)
rel_abundance$id <- rownames(rel_abundance)

tabulated_results_merge <- tabulated_results %>% # create a tabulated results for merging
  select(all_of(c("id","Taxon")))

rel_abundance_taxa <- merge(rel_abundance, tabulated_results_merge, by = "id", all.x = TRUE)

# we now have a table with the per-sample relative abundances of each ASV (along with their taxonomic assignments)

# Taxon column data is messy - let's split the data up into taxonomic ranks
rel_abundance_taxa_split <- rel_abundance_taxa %>% 
  separate(Taxon, 
           into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
           sep = ";\\s*",  # splits on semicolon and optional space
           fill = "right", # fills missing columns with NA if some rows are shorter
           remove = TRUE) # remove original "Taxon" column too

# some ranks have a value even though they are unidentified (e.g., g__, s__)
# let's change those to NAs
rel_abundance_taxa_split[rel_abundance_taxa_split == "d__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "p__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "c__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "o__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "f__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "g__"] <- NA
rel_abundance_taxa_split[rel_abundance_taxa_split == "s__"] <- NA

# create a column with the lowest identified taxonomic level for all ASVs
rel_abundance_taxa_lowest <- rel_abundance_taxa_split %>% 
  rowwise() %>%
  mutate(LowestTaxLevel = {
    tax_levels <- c(Domain, Phylum, Class, Order, Family, Genus, Species)
    last_non_na <- tail(na.omit(tax_levels), 1)
    ifelse(length(last_non_na) > 0, last_non_na, NA)
  }) %>%
  ungroup()

# group by the lowest taxonomic level
rel_abundance_taxa_lowest <- rel_abundance_taxa_lowest %>%
  mutate(LowestTaxLevel = ifelse(is.na(LowestTaxLevel), "Unassigned", LowestTaxLevel)) %>% # convert NAs to "Unassigned"
  group_by(LowestTaxLevel) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# save the file
saveRDS(rel_abundance_taxa_lowest, "taxonomy/rel_abundance_taxa_lowest.RDS")

# group at level 5/Family level
rel_abundance_taxa_family <- rel_abundance_taxa_split %>%
  mutate(Family = ifelse(is.na(Family), "Unassigned", Family)) %>% # convert NAs to "Unassigned"
  group_by(Family) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# save the file
saveRDS(rel_abundance_taxa_family, "taxonomy/rel_abundance_taxa_family.RDS")

# this allows you to have either the lowest-identified level or collapsed taxonomic tables 
# however, these are massive and have >400 samples for this dataset
# because this isn't very conducive to nice taxa bar plots, in the next chunk, we are going to filter to one facility only and make a simple taxa bar plot
```

# Taxa bar plot for STAFS samples

```{r}
# graph at the family level for STAFS samples
# convert to long format
rel_abundance_taxa_family_long <- rel_abundance_taxa_family %>%
  pivot_longer(cols = -Family, names_to = "sample_name", values_to = "relative_abundance")

# assign rare taxa (less than .2 rel abund to other)
rel_abundance_taxa_family_long <- rel_abundance_taxa_family_long %>%
  mutate(rare_taxa = if_else(relative_abundance < 0.2, "Other", Family))

# associate metadata with taxa
metadata <- read_tsv("/Users/dominikusraditya/Documents/ANEQ505_HW/pempek/03_metadata/metadata.txt")
metadata_filt <- metadata[, which((names(metadata) %in% c("sample","sample_type","treatment", "day"))==TRUE)]

rel_abundance_taxa_family_long <- rel_abundance_taxa_family_long %>% 
  left_join(metadata_filt, by = "sample")

# filter out control samples
rel_abundance_taxa_family_long <- rel_abundance_taxa_family_long[rel_abundance_taxa_family_long$sample_type != "control",]

# filter to just one facility (STAFS)
rel_abundance_taxa_family_long_STAFS <- rel_abundance_taxa_family_long[rel_abundance_taxa_family_long$treatment == "pair",]

base_colors <- c(
  "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E",
  "#C06C84", "#A6761D", "#666666", "#7f3b6c", "#355c7d",
  "#b15928", "#9e003f", "#ff7f00", "#8dd3c7", "#bebada",
  "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5",
  "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#a6cee3"
)

# Get unique taxa categories
categories <- unique(rel_abundance_taxa_family_long_STAFS$rare_taxa)

# special colors for 'Other' and 'Unassigned'
special_colors <- c("Other" = "gray", "Unassigned" = "black")
remaining <- setdiff(categories, names(special_colors))

# Assign base colors to remaining taxa
if(length(remaining) > length(base_colors)){
  warning("More taxa than base colors, some colors will be recycled")
}
assigned_colors <- base_colors[seq_along(remaining)]
names(assigned_colors) <- remaining

# Combine special and base colors
colors <- c(special_colors, assigned_colors)

#taxa bar plot, plotted at the Family taxonomic level for STAFS samples
taxa_family_add <- ggplot(rel_abundance_taxa_family_long_STAFS, 
                          aes(x = reorder(sample_name, add_0c), 
                              y = relative_abundance, 
                              fill = rare_taxa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom",
        axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold")) +
  labs(x = "Sample, sorted by increasing Day", 
       y = "Relative Abundance", 
       fill = "Family")

taxa_family_add
ggsave(taxa_family_add, filename = "../05_figures/taxa_family_add.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)

# graph taxa by sample type at the family level for STAFS samples only
rel_abundance_taxa_family_long_pair_ordered <- rel_abundance_taxa_family_long_pair %>%
  mutate(sample_type = fct_relevel(sample_type, "fecal")) %>% 
  arrange(sample_type)

taxa_family_sample_type <- ggplot(rel_abundance_taxa_family_long_pair, aes(x = sample, y = relative_abundance, fill = rare_taxa)) +
  facet_wrap(~sample_type, scales = "free_x") +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(legend.position = "bottom",
        axis.text.x = element_blank()) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Family")
taxa_family_sample_type

ggsave(taxa_family_sample_type, filename = "../05_figures/taxa_family_sample_type.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)
```

# Taxa bar plot for ARF samples at Genus level

```{r}
# group at level 6/Genus level
rel_abundance_taxa_genus <- rel_abundance_taxa_split %>%
  mutate(Genus = ifelse(is.na(Family), "Unassigned", Genus)) %>% # convert NAs to "Unassigned"
  group_by(Genus) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))

# save the file
saveRDS(rel_abundance_taxa_genus, "taxonomy/rel_abundance_taxa_genus.RDS")

# graph at the genus level for ARF samples
# convert to long format
rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus %>%
  pivot_longer(cols = -Genus, names_to = "sample_name", values_to = "relative_abundance")

# assign rare taxa (less than .2 rel abund to other)
rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus_long %>%
  mutate(rare_taxa = if_else(relative_abundance < 0.2, "Other", Genus))

# associate metadata with taxa
metadata <- read_tsv("../03_metadata/metadata.txt")
metadata_filt <- metadata[, which((names(metadata) %in% c("sample_name","sample_type","facility", "add_0c"))==TRUE)]

rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus_long %>% 
  left_join(metadata_filt, by = "sample_name")

# filter out control samples
rel_abundance_taxa_genus_long <- rel_abundance_taxa_genus_long[rel_abundance_taxa_genus_long$sample_type != "control",]

# filter to just one facility (STAFS)
rel_abundance_taxa_genus_long_ARF <- rel_abundance_taxa_genus_long[rel_abundance_taxa_genus_long$facility == "ARF",]

base_colors <- c(
  "#1B9E77", "#D95F02", "#7570B3", "#E7298A", "#66A61E",
  "#C06C84", "#A6761D", "#666666", "#7f3b6c", "#355c7d",
  "#b15928", "#9e003f", "#ff7f00", "#8dd3c7", "#bebada",
  "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5",
  "#d9d9d9", "#bc80bd", "#ccebc5", "#ffed6f", "#a6cee3"
)

# Get unique taxa categories
categories <- unique(rel_abundance_taxa_genus_long_ARF$rare_taxa)

# special colors for 'Other' and 'Unassigned'
special_colors <- c("Other" = "gray", "Unassigned" = "black")
remaining <- setdiff(categories, names(special_colors))

# Assign base colors to remaining taxa
if(length(remaining) > length(base_colors)){
  warning("More taxa than base colors, some colors will be recycled")
}
assigned_colors <- base_colors[seq_along(remaining)]
names(assigned_colors) <- remaining

# Combine special and base colors
colors <- c(special_colors, assigned_colors)

#taxa bar plot, plotted at the Family taxonomic level for STAFS samples
taxa_genus_add_arf <- ggplot(rel_abundance_taxa_genus_long_ARF, 
                          aes(x = reorder(sample_name, add_0c), 
                              y = relative_abundance, 
                              fill = rare_taxa)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom",
        axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold")) +
  labs(x = "Sample, sorted by increasing ADD", 
       y = "Relative Abundance", 
       fill = "Genus")

taxa_genus_add_arf
ggsave(taxa_genus_add_arf, filename = "../05_figures/taxa_genus_add.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)

# graph taxa by sample type at the family level for STAFS samples only
rel_abundance_taxa_genus_long_ARF_ordered <- rel_abundance_taxa_genus_long_ARF %>%
  mutate(sample_type = fct_relevel(sample_type, "skin","soil")) %>% 
  arrange(sample_type)

taxa_genus_sample_type_arf <- ggplot(rel_abundance_taxa_genus_long_ARF, aes(x = sample_name, y = relative_abundance, fill = rare_taxa)) +
  facet_wrap(~sample_type, scales = "free_x") +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors) +
  theme(legend.position = "bottom",
        axis.text.x = element_blank()) +
  labs(x = "Sample", y = "Relative Abundance", fill = "Family")
taxa_genus_sample_type_arf

ggsave(taxa_genus_sample_type_arf, filename = "../05_figures/taxa_genus_sample_type_arf.jpeg", device = "jpeg", dpi = 300, width = 12, height = 8)
```
